---
title: 'STRIVE'
subtitle: "Final Analysis"
author: "Hongxu Wang"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_folding: hide
editor_options: 
  chunk_output_type: console
---
```{r setup univariate analysis pfas, include=FALSE, echo=FALSE}
options(knitr.table.format = "html")
knitr::opts_chunk$set(echo = TRUE,warning=FALSE, message = FALSE, cache=FALSE,collapse=FALSE)
# 
source(fs::path(here::here("!libraries.R")))
source(fs::path(here::here("!directories.R")))
# source(fs::path(here::here("functions.R")))
source(fs::path(here::here("!load_cleaned_data.R")))
```

## Simple linear or logistic regression

**Exposure: normalized pfas**

**Covars (included sig vars found in exposure correlation, also excluding some vars): source (both), sex (both), age_at_enrollment (exposure), sq_average_drink_per_day (exposure)**

*sig vars from exposure correlation:'age_at_enrollment','sex','bmi','smoking','drinking history','drinking frequency','water bottled','water other type','water faucet filter', 'source_UNC','source_DUKE','source_Emory','source_NCSU'
*sig vars from exposure correlation:'sex', 'source','trig_mg_d_l'

**Outcome: natural log ALT, natural log AST, alt_cat1, ast_cat1, alt_cat2, ast_cat2**

*alt_cat1/ast_cat1: Norm: <= 29IU/L(males), <= 19IU/L(females); High: > 29IU/L(males), > 19IU/L(females)

*alt_cat2/ast_cat2: Norm: <= 33IU/L(males), <= 25IU/L(females); High: > 33IU/L(males), > 25IU/L(females)
**Analysis: linear regression/logistic regression**

```{r analysis alt ast endpoints}
## outcome: natural log alt, natural log ast
covars <- c("source", "sex", "age_at_enrollment", "sq_average_drink_per_day")

emerging <- c("pfbs", "pf_pe_a","pf_pe_s", "pf_hp_a")
pfas_name_analysis <- c("pf_hx_s","pfda","pfna","pfos",
                        "pf_hp_a","pfbs", "pfoa","pf_pe_a",
                        "pf_un_a","pf_hp_s","pf_do_a","pf_pe_s")
legacy <- c("pf_hx_s","pfda","pfna","pfos",
            "pfoa","pf_un_a","pf_hp_s",
            "pf_do_a")
pfas_name_log2 <- paste0(pfas_name_analysis,"_", "log2")
emerging_log2 <- paste0(emerging, "_", "log2")
legacy_log2 <- paste0(legacy, "_","log2")

#con_end <-c("log_alt", "log_ast")
con_end <-c("log_alt")
#con_end1 <-c("alt_cat1", "alt_cat2", "ast_cat1", "ast_cat2")
con_end1 <-c("alt_cat2")

res_cont <- epiomics::owas(data,
                      var = con_end,  
                      omics = pfas_name_log2,
                      covars = covars,
                      var_exposure_or_outcome = "outcome",  
                      family = "gaussian",  
                      conf_int = TRUE)

## outcome: alt_cat1, ast_cat1, alt_cat2, ast_cat2
res_cat <- epiomics::owas(data,
                      var = con_end1, 
                      omics = pfas_name_log2,
                      covars = covars,
                      var_exposure_or_outcome = "outcome",  
                      family = "binomial", 
                      conf_int = TRUE,
                      ref_group = "Norm")
res <- res_cont %>% mutate(outcome_type = "continuous") %>%
  bind_rows(res_cat %>% mutate(outcome_type ="categorical"))

# Adding sig variable to indicate the significance and rename PFAS
res1 <- res %>%
  mutate(sig = case_when((p_value < 0.05 & adjusted_pval < 0.05) ~ "Sig.(FDR < 0.05)",
                         p_value < 0.05 ~ "Sig.(p value < 0.05)", 
                         TRUE ~ "Not Sig.")) %>%
  mutate(plot_name = case_when(
    grepl("pfba", feature_name) ~ "PFBA",
    grepl("pf_hp_a", feature_name) ~ "PFHpA",
    grepl("pf_pe_s", feature_name) ~ "PFPeS",
    grepl("pfbs", feature_name) ~ "PFBS",
    grepl("pf_hx_a", feature_name) ~ "PFHxA",
    grepl("pf_do_a", feature_name) ~ "PFDoA",
    grepl("pfda", feature_name) ~ "PFDA",
    grepl("pf_pe_a", feature_name) ~ "PFPeA",
    grepl("pf_un_a", feature_name) ~ "PFUnA",
    grepl("pfna", feature_name) ~ "PFNA",
    grepl("pfos", feature_name) ~ "PFOS",
    grepl("pf_hx_s", feature_name) ~ "PFHxS",
    grepl("pf_hp_s", feature_name) ~ "PFHpS",
    grepl("pfoa", feature_name) ~ "PFOA",
    grepl("pf3ons", feature_name) ~ "9CL-PF3ONS"
  ))%>%
  mutate(group = ifelse(feature_name %in% legacy_log2, "Long-chain", "Short-chain"))

res1$var_name<-factor(res1$var_name, levels=c("log_alt", "log_ast", "alt_cat1", "alt_cat2", "ast_cat1", "ast_cat2"), labels=c("natural log alt", "natural log ast", "lowest cut alt", "highest cut alt", "lowest cut ast", "highest cut ast"))

#write_csv(res1, fs::path(dir_result, "res_single_exposure_analysis_0218.csv"))

```

# cirrhosis in subset
```{r}
## outcome: case_control for cirrhosis
#(dim(data[data$source=='NCSU',])[1]/dim(data)[1])*100
data_ex_ncsu <- data[data$source !='NCSU',]

data_renamed <- data_ex_ncsu[, c('log_ast_alt', 'log_ast', 'log_alt', 'case_control')]
colnames(data_renamed) <- c('natural log AST/ALT', 'natural log AST', 'natural log ALT', 'Cirrhosis diagnosed')

res_cat <- epiomics::owas(data_ex_ncsu,
                      var = "case_control", 
                      omics = pfas_name_log2,
                      covars = covars,
                      var_exposure_or_outcome = "outcome",  
                      family = "binomial", 
                      conf_int = TRUE,
                      ref_group = "Healthy")

# Adding sig variable to indicate the significance and rename PFAS
res2 <- res_cat %>%
  mutate(sig = case_when((p_value < 0.05 & adjusted_pval < 0.05) ~ "Sig.(FDR < 0.05)",
                         p_value < 0.05 ~ "Sig.(p value < 0.05)", 
                         TRUE ~ "Not Sig.")) %>%
  mutate(plot_name = case_when(
    grepl("pfba", feature_name) ~ "PFBA",
    grepl("pf_hp_a", feature_name) ~ "PFHpA",
    grepl("pf_pe_s", feature_name) ~ "PFPeS",
    grepl("pfbs", feature_name) ~ "PFBS",
    grepl("pf_hx_a", feature_name) ~ "PFHxA",
    grepl("pf_do_a", feature_name) ~ "PFDoA",
    grepl("pfda", feature_name) ~ "PFDA",
    grepl("pf_pe_a", feature_name) ~ "PFPeA",
    grepl("pf_un_a", feature_name) ~ "PFUnA",
    grepl("pfna", feature_name) ~ "PFNA",
    grepl("pfos", feature_name) ~ "PFOS",
    grepl("pf_hx_s", feature_name) ~ "PFHxS",
    grepl("pf_hp_s", feature_name) ~ "PFHpS",
    grepl("pfoa", feature_name) ~ "PFOA",
    grepl("pf3ons", feature_name) ~ "9CL-PF3ONS"
  ))%>%
  mutate(group = ifelse(feature_name %in% legacy_log2, "Long-chain", "Short-chain"))


# Combine res1 and res2
res_combined <- bind_rows(res1, res2)

# Save as a single CSV file
write_csv(res_combined, fs::path(dir_result, "res_single_exposure_analysis_0227.csv"))

# Print message for confirmation
message("Combined results saved as res_single_exposure_analysis_0227.csv")

```

#ISEE 2025_2nd paper
```{r}
## outcome: natural log alt, natural log ast
covars <- c("sex", "age_at_enrollment","source")

emerging <- c("pfbs", "pf_pe_a","pf_pe_s", "pf_hp_a")
pfas_name_analysis <- c("pf_hx_s","pfda","pfna","pfos",
                        "pf_hp_a","pfbs", "pfoa","pf_pe_a",
                        "pf_un_a","pf_hp_s","pf_do_a","pf_pe_s"
)
legacy <- c("pf_hx_s","pfda","pfna","pfos",
            "pfoa","pf_un_a","pf_hp_s",
            "pf_do_a")

pfas_name_log2 <- paste0(pfas_name_analysis,"_", "log2")

emerging_log2 <- paste0(emerging, "_", "log2")
legacy_log2 <- paste0(legacy, "_","log2")

#con_end <-c("log_alt", "log_ast")
con_end <-c("log_trig", "log_ast")
#con_end1 <-c("alt_cat1", "alt_cat2", "ast_cat1", "ast_cat2")
con_end1 <-c("trig_flags", "ast_cat2","tg_ast","tg_ast_alt","alt_ast")

res_cont <- epiomics::owas(data,
                      var = con_end,  
                      omics = pfas_name_log2,
                      covars = covars,
                      var_exposure_or_outcome = "outcome",  
                      family = "gaussian",  
                      conf_int = TRUE)

## outcome: alt_cat1, ast_cat1, alt_cat2, ast_cat2
res_cat <- epiomics::owas(data,
                      var = con_end1, 
                      omics = pfas_name_log2,
                      covars = covars,
                      var_exposure_or_outcome = "outcome",  
                      family = "binomial", 
                      conf_int = TRUE,
                      ref_group = "Norm")
res <- res_cont %>% mutate(outcome_type = "continuous") %>%
  bind_rows(res_cat %>% mutate(outcome_type ="categorical"))

# Adding sig variable to indicate the significance and rename PFAS
res1 <- res %>%
  mutate(sig = case_when((p_value < 0.05 & adjusted_pval < 0.05) ~ "Sig.(FDR < 0.05)",
                         p_value < 0.05 ~ "Sig.(p value < 0.05)", 
                         TRUE ~ "Not Sig.")) %>%
  mutate(plot_name = case_when(
    grepl("pfba", feature_name) ~ "PFBA",
    grepl("pf_hp_a", feature_name) ~ "PFHpA",
    grepl("pf_pe_s", feature_name) ~ "PFPeS",
    grepl("pfbs", feature_name) ~ "PFBS",
    grepl("pf_hx_a", feature_name) ~ "PFHxA",
    grepl("pf_do_a", feature_name) ~ "PFDoA",
    grepl("pfda", feature_name) ~ "PFDA",
    grepl("pf_pe_a", feature_name) ~ "PFPeA",
    grepl("pf_un_a", feature_name) ~ "PFUnA",
    grepl("pfna", feature_name) ~ "PFNA",
    grepl("pfos", feature_name) ~ "PFOS",
    grepl("pf_hx_s", feature_name) ~ "PFHxS",
    grepl("pf_hp_s", feature_name) ~ "PFHpS",
    grepl("pfoa", feature_name) ~ "PFOA",
    grepl("pf3ons", feature_name) ~ "9CL-PF3ONS"
  ))%>%
  mutate(group = ifelse(feature_name %in% legacy_log2, "Long-chain", "Short-chain"))

res1$var_name<-factor(res1$var_name, levels=c("log_trig","log_ast", "trig_flags","ast_cat2","tg_ast","tg_ast_alt","alt_ast"), labels=c("natural log tg", "natural log ast", "high vs normal tg", "elevated ast","elevated tg and ast","elevated all","elevated alt and ast"))

write_csv(res1, fs::path(dir_result, "res_single_exposure_analysis_ISEE.csv"))
```

#ISEE 2025_2nd paper by obesity status
```{r}
library(dplyr)
library(tidyr)
library(readr)
library(fs)

# Define variables
covars <- c("sex", "age_at_enrollment", "source")

emerging <- c("pfbs", "pf_pe_a","pf_pe_s", "pf_hp_a")
legacy <- c("pf_hx_s","pfda","pfna","pfos","pfoa","pf_un_a","pf_hp_s","pf_do_a")
pfas_name_analysis <- c(legacy, "pf_hp_a", "pfbs", "pf_pe_a", "pf_pe_s")

pfas_name_log2 <- paste0(pfas_name_analysis, "_log2")
emerging_log2 <- paste0(emerging, "_log2")
legacy_log2 <- paste0(legacy, "_log2")

con_end <- c("log_trig", "log_ast")
con_end1 <- c("trig_flags", "ast_cat2", "tg_ast", "tg_ast_alt", "alt_ast")

# Create an empty result list
all_results <- list()

# Loop over obesity groups
for (group_label in c("Obesity", "Non-obesity")) {
  
  data_group <- data %>% filter(obesity_status == group_label)
  
  # Continuous outcomes
  res_cont <- epiomics::owas(data_group,
                              var = con_end,  
                              omics = pfas_name_log2,
                              covars = covars,
                              var_exposure_or_outcome = "outcome",  
                              family = "gaussian",  
                              conf_int = TRUE)
  
  # Categorical outcomes
  res_cat <- epiomics::owas(data_group,
                             var = con_end1, 
                             omics = pfas_name_log2,
                             covars = covars,
                             var_exposure_or_outcome = "outcome",  
                             family = "binomial", 
                             conf_int = TRUE,
                             ref_group = "Norm")
  
  # Combine
  res <- res_cont %>% mutate(outcome_type = "continuous") %>%
    bind_rows(res_cat %>% mutate(outcome_type = "categorical")) %>%
    mutate(obesity_group = group_label)
  
  all_results[[group_label]] <- res
}

# Combine both strata
res_combined <- bind_rows(all_results)

# Add significance and plot labels
res1 <- res_combined %>%
  mutate(
    sig = case_when(
      p_value < 0.05 & adjusted_pval < 0.05 ~ "Sig.(FDR < 0.05)",
      p_value < 0.05 ~ "Sig.(p value < 0.05)",
      TRUE ~ "Not Sig."
    ),
    plot_name = case_when(
      grepl("pfba", feature_name) ~ "PFBA",
      grepl("pf_hp_a", feature_name) ~ "PFHpA",
      grepl("pf_pe_s", feature_name) ~ "PFPeS",
      grepl("pfbs", feature_name) ~ "PFBS",
      grepl("pf_hx_a", feature_name) ~ "PFHxA",
      grepl("pf_do_a", feature_name) ~ "PFDoA",
      grepl("pfda", feature_name) ~ "PFDA",
      grepl("pf_pe_a", feature_name) ~ "PFPeA",
      grepl("pf_un_a", feature_name) ~ "PFUnA",
      grepl("pfna", feature_name) ~ "PFNA",
      grepl("pfos", feature_name) ~ "PFOS",
      grepl("pf_hx_s", feature_name) ~ "PFHxS",
      grepl("pf_hp_s", feature_name) ~ "PFHpS",
      grepl("pfoa", feature_name) ~ "PFOA",
      grepl("pf3ons", feature_name) ~ "9CL-PF3ONS"
    ),
    group = ifelse(feature_name %in% legacy_log2, "Long-chain", "Short-chain"),
    var_name = factor(var_name,
                      levels = c("log_trig", "log_ast", "trig_flags", "ast_cat2", "tg_ast", "tg_ast_alt", "alt_ast"),
                      labels = c("natural log tg", "natural log ast", "high vs normal tg", "elevated ast",
                                 "elevated tg and ast", "elevated all", "elevated alt and ast"))
  )

# Save result
write_csv(res1, fs::path(dir_result, "res_single_exposure_analysis_ISEE_stratified_obesity.csv"))

```

