---
title: 'STRIVE'
subtitle: "Final Analysis"
author: "Hongxu Wang"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_folding: hide
editor_options: 
  chunk_output_type: console
---
```{r setup univariate analysis cat pfas, include=FALSE, echo=FALSE}
options(knitr.table.format = "html")
knitr::opts_chunk$set(echo = TRUE,warning=FALSE, message = FALSE, cache=FALSE,collapse=FALSE)

source(fs::path(here::here("!libraries.R")))
source(fs::path(here::here("!directories.R")))
# source(fs::path(here::here("functions.R")))
source(fs::path(here::here("!load_cleaned_data.R")))
```

## Analysis1. Logistic regression 
**Outcome: case vs control**

**Exposure: normalized pfas**

**Covars: source, age, sex, race, rural, smoking, alcohol**

**Analysis: logistic regression**

```{r analysis}
covars <- c("source","age_at_enrollment", "sex", "race_final_label","rural", "smoking","sq_drink_alcohol")

res <- epiomics::owas(data,
                          var = "case_control", 
                          omics = pfas_name_scld,
                          covars = covars,
                          var_exposure_or_outcome = "outcome",
                          family = "binomial",
                          conf_int = TRUE,
                          ref_group = "Healthy") %>%
  mutate(odds_ratio = exp(estimate),
         exp_ci_low = exp(conf_low),
         exp_ci_high = exp(conf_high))

# reorder PFAS by estimate
res1 <- res |> 
  mutate(sig = case_when((p_value < 0.05 & adjusted_pval < 0.05) ~
                           "Sig.(FDR < 0.05)",
                          p_value < 0.05 ~ "Sig.(p value < 0.05)", 
                         p_value >= 0.05 ~ "Not Sig.")) %>%
  mutate(plot_name = case_when(grepl("pfba", feature_name) ~ "PFBA",
                               grepl("pf_hp_a", feature_name) ~ "PFHpA",
                               grepl("pf_pe_s", feature_name) ~ "PFPeS",
                               grepl("pfbs", feature_name) ~ "PFBS",
                               grepl("pf_hx_a", feature_name) ~ "PFHxA",
                               grepl("pf_do_a", feature_name) ~ "PFDoA",
                               grepl("pfda", feature_name) ~ "PFDA",
                               grepl("pf_pe_a", feature_name) ~ "PFPeA",
                               grepl("pf_un_a", feature_name) ~ "PFUnA",
                               grepl("pfna", feature_name) ~ "PFNA",
                               grepl("pfos", feature_name) ~ "PFOS",
                               grepl("pf_hx_s", feature_name) ~ "PFHxS",
                               grepl("pf_hp_s", feature_name) ~ "PFHpS",
                               grepl("pfoa", feature_name) ~ "PFOA",
                               grepl("pf3ons", feature_name) ~ "9CL-PF3ONS",))

#writexl::write_xlsx(res1,
 #                   fs::path(dir_result,
  #                           "exposure outcome association.xlsx"))

## Fig. PFAS and outcome
(plot <-  ggplot(res1,
       aes(x = odds_ratio, y = fct_reorder(plot_name, odds_ratio), color = sig)) +
  geom_errorbar(aes(xmin = exp_ci_low, xmax = exp_ci_high),
                width = 0, position = position_dodge(width = 1)) +
  geom_point(size = 1, position = position_dodge(width = 1)) + 
  geom_vline(xintercept = 1, linetype = 2, color = "grey") +
  xlab("Odds Ratio (95% CI)") +
  theme(
        legend.position = "bottom",
        panel.background = element_rect(fill="white"), 
        strip.background = element_rect(fill = "white"),
        strip.text.y = element_text(angle = 0, hjust = 0),
        text = element_text(size = 15),
        axis.title.y = element_blank(),
        axis.line.x = element_line(color = "black"),
        axis.line.y = element_line(color = "black")) +
  ylab("OR (95% CI) per\ndoubling of PFAS") +
  # scale_color_manual(values = c( "black","purple","red")))
  scale_color_manual(values = c( "black","red")))

ggsave(filename = fs::path(dir_figures,"Fig. Effect of each PFAS on case_control status.jpeg"),width = 8,height = 8, dpi = 300, bg = "white")
```

## Same results by group of PFAS (Emerging vs Legacy)
```{r}
# Categorize PFAS into "emerging" and "legacy" based on `pfas_name_scld`
res1 <- res1 %>%
  mutate(group = case_when(
    grepl("pfbs|pf_pe_a|pf_pe_s|pfba|pf_hx_a", feature_name) ~ "Emerging",
    TRUE ~ "Legacy"
  ))

# Create separate plots for "emerging" and "legacy" groups
plot <- ggplot(res1, aes(x = odds_ratio, y = fct_reorder(plot_name, odds_ratio), color = sig)) +
  geom_errorbar(aes(xmin = exp_ci_low, xmax = exp_ci_high),
                width = 0, position = position_dodge(width = 1)) +
  geom_point(size = 1, position = position_dodge(width = 1)) + 
  geom_vline(xintercept = 1, linetype = 2, color = "grey") +
  xlab("Odds Ratio (95% CI)") +
  theme(
    legend.position = "bottom",
    panel.background = element_rect(fill="white"), 
    strip.background = element_rect(fill = "white"),
    strip.text.y = element_text(angle = 0, hjust = 0),
    text = element_text(size = 15),
    axis.title.y = element_blank(),
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black")
  ) +
  ylab("OR (95% CI) per\ndoubling of PFAS") +
  scale_color_manual(values = c("black", "red"))

# Facet by 'group' to show separate panels for "emerging" and "legacy" groups
plot + facet_wrap(~ group, scales = "free_y")

# Save the plot
ggsave(filename = fs::path(dir_figures, "Fig. Effect of each PFAS on case/control status by PFAS group.jpeg"),
       plot = plot + facet_wrap(~ group, scales = "free_y"),
       width = 8,
       height = 8,
       dpi = 300,
       bg = "white")

```


## Analysis2. Linear regression 
**Outcome: AST/ALT**

**Exposure: normalized pfas**

**Covars: source, age, sex, race, rural, smoking, alcohol**

**Analysis: linear regression**

```{r analysis}
covars <- c("source", "age_at_enrollment", "sex", "race_final_label", "rural", "smoking", "sq_drink_alcohol")

# Perform linear regression using epiomics package
res <- epiomics::owas(data,
                      var = "AST/ALT",  # AST/ALT ratio as the outcome variable
                      omics = pfas_name_scld,
                      covars = covars,
                      var_exposure_or_outcome = "outcome",  
                      family = "gaussian",  # Use "gaussian" for linear regression
                      conf_int = TRUE)

# reorder PFAS by estimate
res1 <- res %>%
  mutate(sig = case_when((p_value < 0.05 & adjusted_pval < 0.05) ~ "Sig.(FDR < 0.05)",
                         p_value < 0.05 ~ "Sig.(p value < 0.05)", 
                         TRUE ~ "Not Sig.")) %>%
  mutate(plot_name = case_when(
    grepl("pfba", feature_name) ~ "PFBA",
    grepl("pf_hp_a", feature_name) ~ "PFHpA",
    grepl("pf_pe_s", feature_name) ~ "PFPeS",
    grepl("pfbs", feature_name) ~ "PFBS",
    grepl("pf_hx_a", feature_name) ~ "PFHxA",
    grepl("pf_do_a", feature_name) ~ "PFDoA",
    grepl("pfda", feature_name) ~ "PFDA",
    grepl("pf_pe_a", feature_name) ~ "PFPeA",
    grepl("pf_un_a", feature_name) ~ "PFUnA",
    grepl("pfna", feature_name) ~ "PFNA",
    grepl("pfos", feature_name) ~ "PFOS",
    grepl("pf_hx_s", feature_name) ~ "PFHxS",
    grepl("pf_hp_s", feature_name) ~ "PFHpS",
    grepl("pfoa", feature_name) ~ "PFOA",
    grepl("pf3ons", feature_name) ~ "9CL-PF3ONS"
  ))

# Plot coefficients and confidence intervals
plot <- ggplot(res1, aes(x = estimate, y = fct_reorder(plot_name, estimate), color = sig)) +
  geom_errorbar(aes(xmin = conf_low, xmax = conf_high), width = 0, position = position_dodge(width = 1)) +
  geom_point(size = 1, position = position_dodge(width = 1)) + 
  geom_vline(xintercept = 0, linetype = 2, color = "grey") +  
  labs(
    title = "Effect of each PFAS on AST/ALT Ratio",
    x = "Coefficient (95% CI)", y= "PFAS") +
  theme(
    legend.position = "bottom",
    panel.background = element_rect(fill="white"), 
    strip.background = element_rect(fill = "white"),
    strip.text.y = element_text(angle = 0, hjust = 0),
    text = element_text(size = 15),
    axis.title.y = element_blank(),
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black"),
    plot.title = element_text(size = 15, hjust = 0.5) 
  ) +
  
  scale_color_manual(values = c("black", "red"))

print(plot)

# Save the plot
ggsave(filename = fs::path(dir_figures, "Fig. Effect of each PFAS on AST_ALT Ratio.jpeg"),
       plot = plot,
       width = 8,
       height = 8,
       dpi = 300,
       bg = "white")

```

## Same results by group of PFAS (Emerging vs Legacy)
```{r}
# Reorder PFAS by estimate and add significance and group classification
res1 <- res1 %>%
  mutate(group = case_when(
    grepl("pfbs|pf_pe_a|pf_pe_s|pfba|pf_hx_a", feature_name) ~ "Emerging",
    TRUE ~ "Legacy"
  ))

# Plot coefficients and confidence intervals, with facets for emerging vs legacy
plot <- ggplot(res1, aes(x = estimate, y = fct_reorder(plot_name, estimate), color = sig)) +
  geom_errorbar(aes(xmin = conf_low, xmax = conf_high), width = 0, position = position_dodge(width = 1)) +
  geom_point(size = 1, position = position_dodge(width = 1)) + 
  geom_vline(xintercept = 0, linetype = 2, color = "grey") +  
  labs(
    title = "Effect of each PFAS on AST/ALT Ratio by PFAS group",
    x = "Coefficient (95% CI)", y = "PFAS"
  ) +
  theme(
    legend.position = "bottom",
    panel.background = element_rect(fill = "white"), 
    strip.background = element_rect(fill = "white"),
    strip.text.y = element_text(angle = 0, hjust = 0),
    text = element_text(size = 15),
    axis.title.y = element_blank(),
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black"),
    plot.title = element_text(size = 15, hjust = 0.5)
  ) +
  scale_color_manual(values = c("black", "red")) +
  facet_wrap(~ group, scales = "free_y")  # Facet by group

# Display the plot
print(plot)

# Save the plot
ggsave(
  filename = fs::path(dir_figures, "Fig. Effect of each PFAS on AST_ALT Ratio by Group.jpeg"),
  plot = plot,
  width = 8,
  height = 8,
  dpi = 300,
  bg = "white"
)

```

